---
title: "Introduction to CompStellation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to CompStellation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Overview

The `CompStellation` package provides an easy way to visualize complex microbiome dynamics. It builds on top of `phyloseq`, `microViz`, and `ggplot2` to offer aesthetically pleasing and highly informative "star/radar" plots. These visualizations are particularly useful for seeing differences in microbial composition and relative abundance alongside ordination spaces.

## Setup

```{r setup, eval=FALSE}
library(CompStellation)
library(phyloseq)
library(dplyr)
library(ggplot2)
library(microViz)

# For demonstrations, we will use the GlobalPatterns dataset
data(GlobalPatterns)
# Filter down to speed up example vignettes
# Subset to a smaller physeq for demonstration, fix taxa with microViz
gp <- subset_samples(GlobalPatterns, SampleType %in% c("Feces", "Skin", "Tongue", "Freshwater", "Mock"))
gp <- gp %>% tax_fix()
# Only taxa without zero counts remain
gp <- subset_taxa(gp, taxa_sums(gp) > 0)
```

## Functionality

### 1. `plot_taxa_star()`

This function creates radar/star plots to visualize the relative abundance of specified taxa (or the top N taxa automatically) across different sample groups.

```{r taxa-star, eval=FALSE}
# Plot the top 4 most abundant orders by SampleType
taxa_plot <- plot_taxa_star(
    physeq = gp,
    sample_var = "SampleType",
    taxa_rank = "Order",
    samplecolumn = "X.SampleID",
    error_bar = "SE",
    plot_order = "hclust"
)
print(taxa_plot)
```

### 2. `plot_ordi_star()`

Ordination is a cornerstone of microbiome analysis, but a conventional scatterplot shows only two axes at a time. `plot_ordi_star` performs Principal Coordinate Analysis (PCoA) on a distance matrix (like Bray-Curtis) and summarises each group's position by plotting its centre on the first five axes as the spokes of a star plot, so all five can be compared at once.

Individual samples are not drawn, and this plot carries no taxonomic information — for ordination position alongside taxonomic composition, see `plot_ordi_star_iris()` below.

```{r pcoa-star, eval=FALSE}
# One panel per group
plot_ordi_star(
    physeq = gp,
    sample_var = "SampleType",
    distance = "bray",
    view_type = "separate"
)
# The same summary with all groups overlaid on one radar
pcoa_plot <- plot_ordi_star(
    physeq = gp,
    sample_var = "SampleType",
    distance = "bray",
    view_type = "together"
)
print(pcoa_plot)
```

A few things to keep in mind when reading these:

- The radial axis is **offset**. Ordination scores are negative as well as positive, so every axis is shifted by a constant to make the values plottable on a radius. Radii are therefore not proportional to the scores and are not comparable between axes; the tick labels show the original, unshifted values.
- The **area** of a polygon depends on the order of the spokes, so it changes if the axes are reordered even though no value has changed. Compare spoke by spoke.
- `error_bar = "IQR"` draws the interquartile range of the samples around the group **median**; `error_bar = "SE"` draws one standard error around the group **mean**. They describe different things — spread versus precision — and the title records which was used.
- The plot is **descriptive**. It does not test whether the groups differ; pair it with something like `vegan::adonis2()` on the same distance matrix before making that claim.

### 2b. `plot_ordi_star_iris()` *(experimental)*

This places the same radar at the centre of a circular stacked bar chart, where each segment of the outer ring is one sample's taxonomic composition. The centre and the ring carry different units — ordination score and relative abundance — so read each against its own scale.

```{r pcoa-iris, eval=FALSE}
plot_ordi_star_iris(
    physeq = gp,
    sample_var = "SampleType",
    taxa_rank = "Phylum",
    distance = "bray",
    view_type = "separate"
)
```


### 3. `coreselect()`

`coreselect` compares the core microbiome across groups (it warns above 7, where the plot gets cluttered). It produces a presence/absence dot plot showing which taxa meet your core definition in each sample type.

This needs groups with substantial overlap in taxonomic composition, and where the taxa exclusive to one group's core are of interest. `GlobalPatterns` does not have that structure, so the chunk below simulates a small dataset that does.

Note on the threshold: `percent_samples` accepts either a proportion in (0, 1] or a percentage in (1, 100], and reports which reading it used. A bare `1` means 100%, not 1% — pass `0.01` if you want 1%.

```{r reproduction_data, eval=FALSE}
set.seed(001)
n_samples_per_group <- 20
groups <- c("Control", "TreatmentA", "TreatmentB")
n_samples <- n_samples_per_group * length(groups)
n_taxa <- 25

# Create Taxa Names
taxa_names <- paste0("Taxon_", LETTERS[1:n_taxa])

# OTU Table
# Initialize with zeros
otu_mat <- matrix(0, nrow = n_taxa, ncol = n_samples)
rownames(otu_mat) <- taxa_names
colnames(otu_mat) <- paste0("Sample", 1:n_samples)
taxa_TABLE <- as.matrix(taxa_names)
rownames(taxa_TABLE) <- taxa_names
taxa_tables <- tax_table(taxa_TABLE)
# Assign groups
group_vec <- rep(groups, each = n_samples_per_group)
meta_df <- data.frame(
    SampleID = colnames(otu_mat),
    Group = group_vec,
    row.names = colnames(otu_mat)
)

# Populate OTU table with patterns

for (isValid in 1:n_samples) {
    grp <- group_vec[isValid]

    # Taxon A/M/W: Core in Control
    if (grp == "Control") {
        otu_mat["Taxon_A", isValid] <- 100
        otu_mat["Taxon_M", isValid] <- 10
        otu_mat["Taxon_W", isValid] <- 100
    }

    # Taxon A/M: Conditional presence in Treatments (40% chance)
    if (grp %in% c("TreatmentB", "TreatmentA") && runif(1) < 0.4) {
        otu_mat[c("Taxon_A", "Taxon_M"), isValid] <- 30
    }

    # Taxon B: Core in Control and TreatmentA
    if (grp %in% c("Control", "TreatmentA") && runif(1) < 0.8) {
        otu_mat["Taxon_B", isValid] <- 50
        otu_mat["Taxon_N", isValid] <- 100
        otu_mat["Taxon_O", isValid] <- 10
    }

    # Taxon B: Conditional presence in TreatmentB (30% chance)
    if (grp == "TreatmentB" && runif(1) < 0.3) {
        otu_mat[c("Taxon_B", "Taxon_N", "Taxon_O"), isValid] <- 30
    }

    # Taxon C: Core in TreatmentA
    if (grp == "TreatmentA") {
        otu_mat["Taxon_C", isValid] <- 200
        otu_mat["Taxon_P", isValid] <- 3
        otu_mat["Taxon_Q", isValid] <- 70
    }
    if (grp == "TreatmentB" && runif(1) < 0.3) {
        otu_mat["Taxon_C", isValid] <- 200
        otu_mat["Taxon_P", isValid] <- 3
        otu_mat["Taxon_Q", isValid] <- 70
    }
    # Taxon C: Conditional presence in TreatmentB (30% chance)
    if (grp == "TreatmentB" && runif(1) < 0.3) {
        otu_mat[c("Taxon_C", "Taxon_P", "Taxon_Q"), isValid] <- 20
    }

    # Taxon D: Core in TreatmentB
    if (grp == "TreatmentB") {
        otu_mat["Taxon_D", isValid] <- 500
        otu_mat["Taxon_R", isValid] <- 70
    }
    if (grp == "TreatmentA" && runif(1) < 0.2) {
        otu_mat["Taxon_D", isValid] <- 500
        otu_mat["Taxon_R", isValid] <- 70
    }
    # Taxon E-L: Random presence (20-70%) across all samples
    if (runif(1) < 0.2) otu_mat[c("Taxon_E", "Taxon_S", "Taxon_T", "Taxon_U", "Taxon_V"), isValid] <- 30
    if (runif(1) < 0.3) otu_mat["Taxon_F", isValid] <- 10
    if (runif(1) < 0.4) otu_mat["Taxon_G", isValid] <- 10
    if (runif(1) < 0.6) otu_mat["Taxon_H", isValid] <- 10
    if (runif(1) < 0.7) otu_mat["Taxon_I", isValid] <- 10
    if (runif(1) < 0.5) otu_mat["Taxon_J", isValid] <- 10
    if (runif(1) < 0.5) otu_mat["Taxon_K", isValid] <- 10
    if (runif(1) < 0.5) otu_mat["Taxon_L", isValid] <- 10
}
physeq <- phyloseq(
    otu_table(otu_mat, taxa_are_rows = TRUE), tax_table(taxa_tables),
    sample_data(meta_df)
)
```
```{r coreselect, eval=FALSE}
# Output a core comparison presence-absence plot
core_res <- coreselect(
    physeq = physeq,
    group_var = "Group",
    percent_samples = 80, # 80% prevalence; 0.8 would mean the same
    abundance_threshold = 0
)

# Render the presence/absence plot
print(core_res)

# The core memberships come back with the plot, so they can be fed
# straight into plot_taxa_star() rather than read off the y-axis
attr(core_res, "core_list")
```
### 4. `plot_core_matrix()`

To compare what taxa are uniquely "core" to different groups, `plot_core_matrix` creates a faceted matrix of star plots. Each quadrant isolates the core taxa present _exclusively_ in one group's core compared to the others.

```{r core-matrix, eval=FALSE}
# Compare unique core taxa between groups
coreplot <- plot_core_matrix(
    physeq = physeq,
    group_var = "Group",
    percent_samples = 0.8, # a proportion: 80%
    abundance_threshold = 0,
    taxa_rank = "OTU",
    samplecolumn = "SampleID",
    log_scale = TRUE
)
print(coreplot)

# "Unique" is a hard set difference, so check how close each call was:
# these are the prevalences of every taxon in every group
attr(coreplot, "unique_core")
attr(coreplot, "prevalence")
```

### 5. `plot_concordant_core()`

The complement of `plot_core_matrix()`: instead of the taxa exclusive to one group's core, the taxa shared by *every* evaluated group's core.

```{r concordant-core, eval=FALSE}
concordant <- plot_concordant_core(
    physeq = physeq,
    group_var = "Group",
    percent_samples = 0.8,
    abundance_threshold = 0,
    taxa_rank = "OTU",
    samplecolumn = "SampleID"
)
print(concordant)
attr(concordant, "concordant_core")
```

## A note on sequencing depth

By default the core functions test presence on the abundances as supplied
(`abundance_type = "counts"`). If library sizes are uneven, a taxon registers as
present more often in deeply sequenced samples, so the core partly measures
sequencing effort rather than biology — the functions warn when library sizes
vary more than tenfold. Rarefying beforehand, or passing
`abundance_type = "relative"` with a non-zero `abundance_threshold`, avoids this.
